---
title: "DGE_DESeq2_master"
author: "Payel Sardar"
format: pdf
---

## Setting the working directory
```{r}
# Change the path to the home directory of the project
setwd("D:/KCL2024/Courses/Research_Project/NanoSeq/")
```

## Installing any missing package
```{r}
#BiocManager::install("apeglm")
```

### Importing packages
```{r}
library(DESeq2)
library(apeglm)
library(tidyverse)
library(ggplot2)
library(plotly)
library(dplyr)
library(vsn)
library(ComplexHeatmap)
library(biomaRt)
```

## Setting the output path
```{r}
# This creates an output directory to store the results from the analysis
output_path <- '../DGE_lf/'  # Alter as needed
dir.create(output_path, recursive = TRUE)
```

# Data Preprocessing

Reading the counts file
```{r}
# Change the path to the counts data as required
read_counts <- read.csv('../Counts/counts_gene_master.txt', header = TRUE, sep = '\t')
head(read_counts)
```

Checking the dimension of the counts data
```{r}
dim(read_counts)
```

Checking if all the gene ids are unique
```{r}
# This returns TRUE if any duplicate rows (gene ids) exists, FALSE otherwise
any(duplicated(read_counts$GENEID))
```
Observation: The gene expression data has no duplicate rows and we can proceed with the Differential Gene Expression Analysis

Removing the gene_id column from our dataframe as DESeq2 expects a dataframe containing only count data
```{r}
rownames(read_counts) <- read_counts$GENEID
read_counts[,1] <- NULL
head(read_counts)

# Round the counts to integers
read_counts <- round(read_counts)
```

# Generating annotations

```{r}
# Creating an ensembl object
ensembl <- useEnsembl(biomart = "genes")

# For selecting the dataset
datasets <- listDatasets(ensembl)
head(datasets)
```

```{r}
# Based on the version of Reference genome used for the alignment and quantification, the following dataset is selected
ensembl_mouse <- useEnsembl(biomart = 'genes', 
                       dataset = 'mmusculus_gene_ensembl',
                       mirror = 'www')
```

```{r}
gene_list <- unique(row.names(read_counts))
```

# Extracting the annotation for the specific set of genes
```{r}
annotations <- getBM(
  attributes = c('ensembl_gene_id', 'external_gene_name', 'description'),
  filters = 'ensembl_gene_id',
  values = gene_list,
  mart = ensembl_mouse
)
```

Viewing the first few rows of the annotations dataframe
```{r}
head(annotations)
```

```{r}
# Saving the annotations for any downstream analysis
write.csv(annotations,
          "../metadata/NanoSeq_gene_annotations_ensembl.csv", 
          row.names = FALSE)
```

# Creating the design matrix for DGE
```{r}

design_df <- data.frame(
  sample_id = colnames(read_counts),
  stage = sub("_.*", "", colnames(read_counts)),
  condition = sub("^[^_]*_([^_]*).*", "\\1", colnames(read_counts))
)

# Fix stage name
design_df$stage <- gsub("E18", "E18.5", design_df$stage)

# Create the group column
design_df$group <- paste(design_df$stage, design_df$condition, sep = "_")

# Reorder if needed
design_df <- design_df[, c("sample_id","stage", "condition", "group")]

design_df
```

Re-ordering the columns of gene counts to match the order of sample_id in design matrix. It can be done the other way around as well
```{r}
read_counts <- read_counts[, match(design_df$sample_id, colnames(read_counts))]
```

```{r}
zero_rows <- apply(read_counts, 1, \(x) all(x == 0))
```

```{r}
# Removing rows (genes) where counts = 0 across all samples
read_counts_2 <- read_counts[!zero_rows, ]
```


## Creating a summarized experiment object
```{r}
stopifnot(design_df$sample_id == colnames(read_counts_2)) # samples

se <- SummarizedExperiment(
    assays = list(counts = as.matrix(read_counts_2)),
    colData = design_df)
```

### Access the counts
```{r}
head(assay(se),3)
```

### Access the sample annotations
```{r}
colData(se)
```


```{r}
# Removing genes/rows that do not have > 5 total counts (Removing Noise)
se <- se[rowSums(assay(se, "counts")) > 5, ]
```

## Visualising the distribution of Library Size across all samples
```{r}
# Add in the sum of all counts

se$libSize <-  colSums(assay(se))

colData(se) |>
  as.data.frame() |>
  mutate(sample_id = factor(sample_id, levels = sample_id[order(group)])) |>
  ggplot(aes(x = sample_id, y = libSize / 1e6, fill = group)) + 
    geom_bar(stat = "identity") + 
    theme_bw() + 
    labs(x = "Sample", y = "Total count in millions") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
    scale_x_discrete(labels = function(x) sub("\\.sorted$", "", x))

```

```{r}
# Saving the plot
ggsave("../Plots/NanoSeq_libSize.png", width = 8, height = 6, dpi = 300)
```

## Adjusting for differences in library size
```{r}
dds <- DESeq2::DESeqDataSet(se, design = ~ stage + condition)
```

```{r}
dds <- estimateSizeFactors(dds)

ggplot(data.frame(libSize = colSums(assay(dds)),
                  sizeFactor = sizeFactors(dds),
                  Group = dds$group),
       aes(x = libSize, y = sizeFactor, col = Group)) + 
    geom_point(size = 5) + theme_bw() + 
    labs(x = "Library size", y = "Size factor")
```

```{r}
# Saving the plot
ggsave("../Plots/NanoSeq_SizeFactor.png", width = 8, height = 6, dpi = 300)
```

## Is the variance dependent on the mean?
```{r}
p <- meanSdPlot(assay(dds), ranks = FALSE, plot = FALSE)
p$gg + 
  ggtitle("Before Variance Stabilisation") +
  theme(plot.title = element_text(hjust = 0.5))
```

It is observed that the variance increases with the average read count.

### Variance stabilisation
Transforming the data using DESeq2’s variance stabilizing transformation and then verifying if it has removed the correlation between average read count and variance.

```{r}
vsd <- DESeq2::vst(dds, blind = FALSE)
p <- meanSdPlot(assay(vsd), ranks = FALSE)
p$gg + 
  ggtitle("After Variance Stabilisation") +
  theme(plot.title = element_text(hjust = 0.5))
```

# Saving the variance stabilised normalised expression values
```{r}
normalized_counts <- assay(vsd)
head(normalized_counts)
write.csv(normalized_counts, '../DGE/vsd_norm_gene_counts_NanoSeq.csv', row.names = TRUE)
```

```{r}
long_df <- normalized_counts %>%
     as.data.frame() %>%
     rownames_to_column("gene_id") %>%
     pivot_longer(
         cols = -gene_id,
         names_to = "sample_id",
         values_to = "expression" 
         )
merged_df <- long_df %>%
    inner_join(design_df, by = "sample_id")

agg_df <- merged_df %>%
     group_by(gene_id, stage, condition) %>%
     summarise(
         mean_expr = mean(expression),
         sd_expr = sd(expression),
         .groups = "drop"
     ) %>%
     mutate(stage = factor(stage, levels = c("E18.5", "P1", "P3")))

agg_df <- agg_df %>%
     left_join(annotations %>% dplyr::select(ensembl_gene_id, external_gene_name), 
               by = c("gene_id" = "ensembl_gene_id"))

write.csv(agg_df, '../DGE/Grouped_norm_gene_counts_NanoSeq.csv',row.names = FALSE)
```


## Clustering samples based on their similarity of expression patterns (Euclidean distance)
```{r}
# Define colors for the heatmap
colors <- colorRampPalette(c("blue", "white", "red"))(255)

# Compute Euclidean distances between samples
dst <- dist(t(assay(vsd)))  # transpose to get sample-wise distance

# Create the heatmap with annotations and values displayed
ComplexHeatmap::Heatmap(
    as.matrix(dst), 
    col = colors,
    name = "Euclidean\ndistance",
    cluster_rows = hclust(dst),
    cluster_columns = hclust(dst),
    bottom_annotation = HeatmapAnnotation(
        condition = vsd$condition,
        stage = vsd$stage,
        col = list(
            condition = c(WT = "orange", Mut = "pink"),
            stage = c(E18.5 = "yellow", P1 = "forestgreen", P3 = "purple")
        )
    )
)

```

## PCA Plot

```{r}
pcaData <- DESeq2::plotPCA(vsd, intgroup = c("condition", "stage"),
                           returnData = TRUE)
```

Making the PCA plot
```{r}
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(x = PC1, y = PC2)) +
    geom_point(aes(color = condition, shape = stage), size = 3) +
    theme_minimal() +
    xlab(paste0("PC1: ", percentVar[1], "% variance")) +
    ylab(paste0("PC2: ", percentVar[2], "% variance")) +
    coord_fixed() + 
    scale_color_manual(values = c(WT = "black", Mut = "darkmagenta"))+
    scale_shape_manual(values = c(15,16,17)) +
  ggtitle("Principal Component Analysis (PCA)")
```

# Differential Expression Analysis

Normalization
```{r}
dds <- estimateSizeFactors(dds)
```

Statistical Modelling
```{r}
dds <- estimateDispersions(dds)
```

```{r}
plotDispEsts(dds)
title("Dispersion Estimates Across Mean Expression")
```

Testing
```{r}
dds <- nbinomWaldTest(dds)
```

```{r}
levels(dds$stage)
```
```{r}
# Relevel condition factor to set WT as reference
colData(dds)$condition <- relevel(colData(dds)$condition, ref = "WT")

# Verify new level order
levels(colData(dds)$condition)
```


### Function to perform DGE based on stage and condition as per requirement

```{r}
run_DGE_analysis <- function(dds, group_by = "condition", group_value, contrast_var, contrast, annotations, 
                             log2FC_threshold = 1.2, padj_threshold = 0.05, 
                             output_prefix = "../DGE_lf",
                             file_tag = "Annotated_sig_DGE") {
  # Subset DDS object
  dds_sub <- dds[, colData(dds)[[group_by]] == group_value]
  
  # Drop unused factor levels in the contrast variable
  colData(dds_sub)[[contrast_var]] <- droplevels(factor(colData(dds_sub)[[contrast_var]]))
  
  # Set reference level for contrast variable to the SECOND element of contrast
  colData(dds_sub)[[contrast_var]] <- relevel(
    colData(dds_sub)[[contrast_var]], 
    ref = contrast[2]
  )
  
  # Set design
  design(dds_sub) <- as.formula(paste("~", contrast_var))
  
  # Run DESeq
  dds_sub <- DESeq2::DESeq(dds_sub)
  
  # Perform contrast with log2FC shrinkage
  coef_name <- resultsNames(dds_sub)[grep(paste0(contrast_var, "_", contrast[1], "_vs_", contrast[2]), resultsNames(dds_sub))]
  if (length(coef_name) != 1) stop("Contrast coefficient not found or ambiguous.")
  
  res <- lfcShrink(dds_sub, coef = coef_name, type = "apeglm")
  summary(res)
  
  # Convert to data frame
  res_df <- as.data.frame(res)
  res_df <- res_df[order(res_df$log2FoldChange, decreasing = TRUE), ]
  
  # Write raw results
  raw_outfile <- file.path(output_prefix, paste0(file_tag, "_", group_by, "_", group_value, "_", contrast[1], "_vs_", contrast[2], ".csv"))
  write.csv(res_df, raw_outfile, row.names = TRUE)
  
  # Filter significant genes
  sig_res <- subset(res_df, padj <= padj_threshold & abs(log2FoldChange) >= log2FC_threshold)
  sig_res$gene_id <- rownames(sig_res)
  
  # Annotate
  annotated_res <- merge(sig_res, annotations, by.x = "gene_id", by.y = "ensembl_gene_id", all.x = TRUE)
  
  # Write annotated significant DEGs with customizable tag
  annotated_outfile <- file.path(output_prefix, paste0(file_tag, "_", "Annotated_sig" , "_", group_by, "_", group_value, "_", contrast[1], "_vs_", contrast[2], ".csv"))
  write.csv(annotated_res[order(annotated_res$log2FoldChange, decreasing = TRUE), ],
            annotated_outfile, row.names = FALSE)
  
  return(list(dds = dds_sub, results = res, sig_results = annotated_res))
}
```

# For Stage Wise DGE

## For E18.5

```{r}
res_E18.5 <- run_DGE_analysis(
  dds = dds,
  group_by = "stage",
  group_value = "E18.5",
  contrast_var = "condition",
  contrast = c("Mut", "WT"),
  annotations = annotations,
  file_tag = "StageWiseDGE"
)
```
## For P1
```{r}
res_P1 <- run_DGE_analysis(
  dds = dds,
  group_by = "stage",
  group_value = "P1",
  contrast_var = "condition",
  contrast = c("Mut", "WT"),
  annotations = annotations,
  file_tag = "StageWiseDGE"
)
```

## For P3
```{r}
res_P3 <- run_DGE_analysis(
  dds = dds,
  group_by = "stage",
  group_value = "P3",
  contrast_var = "condition",
  contrast = c("Mut", "WT"),
  annotations = annotations,
  file_tag = "StageWiseDGE"
)
```
# Temporal DGE Analysis

## For Wild Type

### For E18.5 vs P1
```{r}
res_WT_E18.5vsP1 <- run_DGE_analysis(
  dds = dds,
  group_by = "condition",
  group_value = "WT",
  contrast_var = "stage",
  contrast = c("P1", "E18.5"),
  annotations = annotations,
  file_tag = "TemporalDGE"
)
```
### For P1 vs P3

```{r}
res_WT_P1vsP3 <- run_DGE_analysis(
  dds = dds,
  group_by = "condition",
  group_value = "WT",
  contrast_var = "stage",
  contrast = c("P3", "P1"),
  annotations = annotations,
  file_tag = "TemporalDGE"
)
```

### For E18.5 vs P3
```{r}
res_WT_E18.5vsP3 <- run_DGE_analysis(
  dds = dds,
  group_by = "condition",
  group_value = "WT",
  contrast_var = "stage",
  contrast = c("P3", "E18.5"),
  annotations = annotations,
  file_tag = "TemporalDGE"
)
```
## For Mutant

### For E18.5 vs P1

```{r}
res_Mut_E18.5vsP1 <- run_DGE_analysis(
  dds = dds,
  group_by = "condition",
  group_value = "Mut",
  contrast_var = "stage",
  contrast = c("P1", "E18.5"),
  annotations = annotations,
  file_tag = "TemporalDGE"
)
```

### For P1 vs P3

```{r}
res_Mut_P1vsP3 <- run_DGE_analysis(
  dds = dds,
  group_by = "condition",
  group_value = "Mut",
  contrast_var = "stage",
  contrast = c("P3", "P1"),
  annotations = annotations,
  file_tag = "TemporalDGE"
)
```

### For E18.5 vs P3

```{r}
res_Mut_E18.5vsP3 <- run_DGE_analysis(
  dds = dds,
  group_by = "condition",
  group_value = "Mut",
  contrast_var = "stage",
  contrast = c("P3", "E18.5"),
  annotations = annotations,
  file_tag = "TemporalDGE"
)
```




