---
title: "DTE_DESeq2_master"
author: "Payel Sardar"
format: pdf
---

## Setting the working directory
```{r}
setwd("D:/KCL2024/Courses/Research_Project/NanoSeq/")
```

## Installing any missing package
```{r}
#BiocManager::install("apeglm")
```

### Importing packages
```{r}
library(DESeq2)
library(apeglm)
library(tidyverse)
library(ggplot2)
library(plotly)
library(dplyr)
library(vsn)
library(ComplexHeatmap)
library(biomaRt)
```

## Setting the output path
```{r}
output_path <- '../DTE/'
#dir.create(output_path, recursive = TRUE)
```

# Data Preprocessing

Reading the counts file
```{r}
read_counts <- read.csv('../Counts/counts_transcript_master.txt', header = TRUE, sep = '\t')
head(read_counts)
```

Checking the dimension of the counts data
```{r}
dim(read_counts)
```

Checking if all the transcript ids are unique
```{r}
# This returns TRUE if any duplicate rows (transcript ids) exists, FALSE otherwise
any(duplicated(read_counts$TXNAME))
```
Observation: The transcript quantification data has no duplicate rows and we can proceed with the Differential transcript Expression Analysis

## Extracting the first 2 columns to preserve the transcript-gene mapping
```{r}
tx_gene_mapping <- read_counts[,c(1,2)]
head(tx_gene_mapping)

# Dropping the gene id column from the reads data
read_counts$GENEID <- NULL
```


Removing the transcript id (TXNAME) column from our dataframe as DESeq2 expects a dataframe containing only count data
```{r}
rownames(read_counts) <- read_counts$TXNAME
read_counts[,1] <- NULL

# Round the counts to integers
read_counts <- round(read_counts)

head(read_counts)
```

# Generating annotations

```{r}
# Creating an ensembl object
ensembl <- useEnsembl(biomart = "genes")

# For selecting the dataset
datasets <- listDatasets(ensembl)
head(datasets)
```

```{r}
# Based on the version of Reference genome used for the alignment and quantification, the following dataset is selected
ensembl_mouse <- useEnsembl(biomart = 'genes', 
                       dataset = 'mmusculus_gene_ensembl',
                       mirror = 'www')
```

```{r}
tx_list <- unique(row.names(read_counts))
```

# Extracting the annotation for the specific set of genes
```{r}
annot_tx <- getBM(
  attributes = c('ensembl_transcript_id', 'ensembl_gene_id',
                 'external_gene_name', 'transcript_biotype', 
                 'chromosome_name', 'start_position', 'end_position', 
                 'strand','description'),
  filters = 'ensembl_transcript_id',
  values = tx_list,
  mart = ensembl_mouse
)
head(annot_tx)
```
Merge transcript-level annotation with tx_gene_mapping
```{r}
annot_all <- merge(tx_gene_mapping, annot_tx,
                   by.x = "TXNAME", by.y = "ensembl_transcript_id",
                   all.x = TRUE)
```

 Identify transcripts with missing transcript-level annotations
```{r}
missing_tx <- annot_all[is.na(annot_all$external_gene_name), ]
missing_gene_ids <- unique(missing_tx$GENEID)
```

Fetch gene-level annotations for fallback
```{r}
annot_gene <- getBM(
  attributes = c('ensembl_gene_id', 'external_gene_name', 'description'),
  filters = 'ensembl_gene_id',
  values = missing_gene_ids,
  mart = ensembl_mouse
)
```

Merge gene-level annotations into missing entries
```{r}
annot_all <- merge(annot_all, annot_gene,
                   by.x = "GENEID", by.y = "ensembl_gene_id",
                   all.x = TRUE, suffixes = c("", "_gene"))
```

Fill in missing transcript-level fields from gene-level annotation
```{r}
annot_all$external_gene_name <- ifelse(is.na(annot_all$external_gene_name),
                                       annot_all$external_gene_name_gene,
                                       annot_all$external_gene_name)

annot_all$description <- ifelse(is.na(annot_all$description),
                                annot_all$description_gene,
                                annot_all$description)

# Clean up helper columns
annot_all <- annot_all[, !grepl("_gene$", names(annot_all))]

tail(annot_all)
```


```{r}
# Saving the annotations for any downstream analysis
write.csv(annot_all,
          "../metadata/NanoSeq_transcript_annotations_ensembl.csv", 
          row.names = FALSE)
```

# Creating the design matrix for DTE
```{r}

design_df <- data.frame(
  sample_id = colnames(read_counts),
  stage = sub("_.*", "", colnames(read_counts)),
  condition = sub("^[^_]*_([^_]*).*", "\\1", colnames(read_counts))
)

# Fix stage name
design_df$stage <- gsub("E18", "E18.5", design_df$stage)

# Create the group column
design_df$group <- paste(design_df$stage, design_df$condition, sep = "_")

# Reorder if needed
design_df <- design_df[, c("sample_id","stage", "condition", "group")]

design_df
```

Re-ordering the columns of gene counts to match the order of sample_id in design matrix. It can be done the other way around as well
```{r}
read_counts <- read_counts[, match(design_df$sample_id, colnames(read_counts))]
```

```{r}
zero_rows <- apply(read_counts, 1, \(x) all(x == 0))
table(zero_rows)
```

```{r}
# Removing rows (genes) where counts = 0 across all samples
read_counts_2 <- read_counts[!zero_rows, ]
dim(read_counts_2)
```


## Creating a summarized experiment object
```{r}
stopifnot(design_df$sample_id == colnames(read_counts_2)) # samples

se <- SummarizedExperiment(
    assays = list(counts = as.matrix(read_counts_2)),
    colData = design_df)
```

### Access the counts
```{r}
head(assay(se))
```

### Access the sample annotations
```{r}
colData(se)
```


```{r}
# Removing transcripts(rows) that do not have > 5 total counts 
se <- se[rowSums(assay(se, "counts")) > 5, ]
nrow(se)
```

## Visualising the distribution of Library Size across all samples
```{r}
# Add in the sum of all counts

se$libSize <-  colSums(assay(se))

colData(se) |>
  as.data.frame() |>
  mutate(sample_id = factor(sample_id, levels = sample_id[order(group)])) |>
  ggplot(aes(x = sample_id, y = libSize / 1e6, fill = group)) + 
         geom_bar(stat = "identity") + theme_bw() + 
         labs(x = "Sample", y = "Total count in millions") + 
         theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1))
```

## Adjusting for differences in library size
```{r}
dds <- DESeq2::DESeqDataSet(se, design = ~ stage + condition)
```

```{r}
dds <- estimateSizeFactors(dds)

ggplot(data.frame(libSize = colSums(assay(dds)),
                  sizeFactor = sizeFactors(dds),
                  Group = dds$group),
       aes(x = libSize, y = sizeFactor, col = Group)) + 
    geom_point(size = 5) + theme_bw() + 
    labs(x = "Library size", y = "Size factor")
```

## Is the variance dependent on the mean?
```{r}
meanSdPlot(assay(dds), ranks = FALSE)
```

It is observed that the variance increases with the average read count.

### Variance stabilisation
Transforming the data using DESeq2’s variance stabilizing transformation and then verifying if it has removed the correlation between average read count and variance.

```{r}
vsd <- DESeq2::vst(dds, blind = TRUE)
meanSdPlot(assay(vsd), ranks = FALSE)
```

## Clustering samples based on their similarity of expression patterns (Euclidean distance)
```{r}
# Define colors for the heatmap
colors <- colorRampPalette(c("blue", "white", "red"))(255)

# Compute Euclidean distances between samples
dst <- dist(t(assay(vsd)))  # transpose to get sample-wise distance

# Create the heatmap with annotations and values displayed
ComplexHeatmap::Heatmap(
    as.matrix(dst), 
    col = colors,
    name = "Euclidean\ndistance",
    cluster_rows = hclust(dst),
    cluster_columns = hclust(dst),
    bottom_annotation = HeatmapAnnotation(
        condition = vsd$condition,
        stage = vsd$stage,
        col = list(
            condition = c(WT = "orange", Mut = "pink"),
            stage = c(E18.5 = "yellow", P1 = "forestgreen", P3 = "purple")
        )
    )
)

```

## PCA Plot

```{r}
pcaData <- DESeq2::plotPCA(vsd, intgroup = c("condition", "stage"),
                           returnData = TRUE)
```


```{r}
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(x = PC1, y = PC2)) +
    geom_point(aes(color = condition, shape = stage), size = 3) +
    theme_minimal() +
    xlab(paste0("PC1: ", percentVar[1], "% variance")) +
    ylab(paste0("PC2: ", percentVar[2], "% variance")) +
    coord_fixed() + 
    scale_color_manual(values = c(WT = "black", Mut = "darkmagenta"))+
    scale_shape_manual(values = c(15,16,17))
```

# Differential Expression Analysis

Normalization
```{r}
dds <- estimateSizeFactors(dds)
```

Statistical Modelling
```{r}
dds <- estimateDispersions(dds)
```

```{r}
plotDispEsts(dds)
```

Testing
```{r}
dds <- nbinomWaldTest(dds)
```

```{r}
levels(dds$stage)
```

### Function to perform DTE based on stage and condition as per requirement

```{r}
run_DTE_analysis <- function(dds, group_by = "condition", group_value, contrast_var, contrast, 
                             annotations_tx, tx2gene,
                             log2FC_threshold = 1.2, padj_threshold = 0.05, 
                             output_prefix = "../DTE",
                             file_tag = "Annotated_sig_DTE") {
  # Subset DDS object by group
  dds_sub <- dds[, colData(dds)[[group_by]] == group_value]
  
  # Drop unused factor levels in contrast variable
  colData(dds_sub)[[contrast_var]] <- droplevels(factor(colData(dds_sub)[[contrast_var]]))
  
  # Set design formula
  design(dds_sub) <- as.formula(paste("~", contrast_var))
  
  # Run DESeq
  dds_sub <- DESeq2::DESeq(dds_sub)
  
  # Extract contrast results
  res <- results(dds_sub, contrast = c(contrast_var, contrast[1], contrast[2]))
  summary(res)
  
  # Prepare results dataframe
  res_df <- as.data.frame(res)
  res_df$transcript_id <- rownames(res_df)
  res_df <- res_df[order(res_df$log2FoldChange, decreasing = TRUE), ]
  
  # Write raw results to file
  raw_outfile <- file.path(output_prefix, paste0(file_tag, "_", group_by, "_", group_value, "_", contrast[1], "_vs_", contrast[2], ".csv"))
  write.csv(res_df, raw_outfile, row.names = FALSE)
  
  # Filter significant transcripts
  sig_res <- subset(res_df, padj <= padj_threshold & abs(log2FoldChange) >= log2FC_threshold)
  
  # Merge with transcript-level annotation
  annotated_res <- merge(sig_res, annotations_tx, by.x = "transcript_id", by.y = "TXNAME", all.x = TRUE)
  
  # Handle missing gene annotations (fallback to gene-level using tx2gene)
  missing <- is.na(annotated_res$external_gene_name)
  
  if (any(missing)) {
    # Map transcript to gene ID using tx2gene
    tx2gene_match <- tx2gene[match(annotated_res$transcript_id[missing], tx2gene$TXNAME), ]
    annotated_res$GENEID[missing] <- tx2gene_match$GENEID

    # Query missing gene annotations using biomaRt
    missing_gene_ids <- unique(annotated_res$GENEID[missing])
    annot_gene <- getBM(
      attributes = c('ensembl_gene_id', 'external_gene_name', 'description'),
      filters = 'ensembl_gene_id',
      values = missing_gene_ids,
      mart = useEnsembl(biomart = "genes", dataset = "mmusculus_gene_ensembl", mirror = "www")
    )

    # Merge with gene-level annotation
    annotated_res <- merge(annotated_res, annot_gene, by.x = "GENEID", by.y = "ensembl_gene_id", all.x = TRUE, suffixes = c("", "_gene"))

    # Fill in missing fields
    annotated_res$external_gene_name <- ifelse(is.na(annotated_res$external_gene_name),
                                               annotated_res$external_gene_name_gene,
                                               annotated_res$external_gene_name)
    annotated_res$description <- ifelse(is.na(annotated_res$description),
                                        annotated_res$description_gene,
                                        annotated_res$description)

    # Remove helper columns
    annotated_res <- annotated_res[, !grepl("_gene$", names(annotated_res))]
  }
  
  # Write annotated results to file
  annotated_outfile <- file.path(output_prefix, paste0(file_tag, "_", "Annotated_sig", "_", group_by, "_", group_value, "_", contrast[1], "_vs_", contrast[2], ".csv"))
  write.csv(annotated_res[order(annotated_res$log2FoldChange, decreasing = TRUE), ],
            annotated_outfile, row.names = FALSE)
  
  return(list(dds = dds_sub, results = res, sig_results = annotated_res))
}

```

# For Stage Wise DGE

## For E18.5

```{r}
res_E18.5 <- run_DTE_analysis(
  dds = dds,
  group_by = "stage",
  group_value = "E18.5",
  contrast_var = "condition",
  contrast = c("Mut", "WT"),
  annotations_tx = annot_all,
  tx2gene = tx_gene_mapping,
  file_tag = "StageWiseDTE"
)
```

## For P1
```{r}
res_P1 <- run_DTE_analysis(
  dds = dds,
  group_by = "stage",
  group_value = "P1",
  contrast_var = "condition",
  contrast = c("Mut", "WT"),
  annotations_tx = annot_all,
  tx2gene = tx_gene_mapping,
  file_tag = "StageWiseDTE"
)
```

## For P3
```{r}
res_P3 <- run_DTE_analysis(
  dds = dds,
  group_by = "stage",
  group_value = "P3",
  contrast_var = "condition",
  contrast = c("Mut", "WT"),
  annotations_tx = annot_all,
  tx2gene = tx_gene_mapping,
  file_tag = "StageWiseDTE"
)
```

# Temporal DGE Analysis

## For Wild Type

### For E18.5 vs P1
```{r}
res_WT_E18.5vsP1 <- run_DTE_analysis(
  dds = dds,
  group_by = "condition",
  group_value = "WT",
  contrast_var = "stage",
  contrast = c("P1", "E18.5"),
  annotations_tx = annot_all,
  tx2gene = tx_gene_mapping,
  file_tag = "TemporalDTE"
)
```

### For P1 vs P3

```{r}
res_WT_P1vsP3 <- run_DTE_analysis(
  dds = dds,
  group_by = "condition",
  group_value = "WT",
  contrast_var = "stage",
  contrast = c("P3", "P1"),
  annotations_tx = annot_all,
  tx2gene = tx_gene_mapping,
  file_tag = "TemporalDTE"
)
```
### For E18.5 vs P3
```{r}
res_WT_E18.5vsP3 <- run_DTE_analysis(
  dds = dds,
  group_by = "condition",
  group_value = "WT",
  contrast_var = "stage",
  contrast = c("P3", "E18.5"),
  annotations_tx = annot_all,
  tx2gene = tx_gene_mapping,
  file_tag = "TemporalDTE"
)
```

## For Mutant

### For E18.5 vs P1

```{r}
res_Mut_E18.5vsP1 <- run_DTE_analysis(
  dds = dds,
  group_by = "condition",
  group_value = "Mut",
  contrast_var = "stage",
  contrast = c("P1", "E18.5"),
  annotations_tx = annot_all,
  tx2gene = tx_gene_mapping,
  file_tag = "TemporalDTE"
)
```

### For P1 vs P3

```{r}
res_Mut_P1vsP3 <- run_DTE_analysis(
  dds = dds,
  group_by = "condition",
  group_value = "Mut",
  contrast_var = "stage",
  contrast = c("P3", "P1"),
  annotations_tx = annot_all,
  tx2gene = tx_gene_mapping,
  file_tag = "TemporalDTE"
)
```

### For E18.5 vs P3

```{r}
res_Mut_E18.5vsP3 <- run_DTE_analysis(
  dds = dds,
  group_by = "condition",
  group_value = "Mut",
  contrast_var = "stage",
  contrast = c("P3", "E18.5"),
  annotations_tx = annot_all,
  tx2gene = tx_gene_mapping,
  file_tag = "TemporalDTE"
)
```




